---
title: "Volleyball - Women's NCAA Division I"
author: "YOUR NAME HERE"
format:
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: tango
editor: visual
editor_options:
  chunk_output_type: console
---

# Getting started: Volleyball data

```{r, message = FALSE}
library(tidyverse) #loads package
library(ggridges)

volleyball <- read_csv("volleyball_ncaa_div1_2022_23_clean.csv") #loads data
```

```{r}
glimpse(volleyball)
```

## Exercise 1

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

## Exercise 2

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

f.  YOUR ANSWER HERE

# One numeric variable

## Exercise 3

a.  

```{r}
# your code here - remove the hashtag! 
```

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

## Exercise 4

a.  

```{r}
#your code here - remove the hashtag! 
```

b.  YOUR ANSWER HERE

c.  

```{r}
#your code here - remove the hashtag! 
```

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

## Exercise 5

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

## Exercise 6

a.  

```{r}
# your code here - remove the hashtag! 
```

b.  YOUR ANSWER HERE

# Two numeric variables

## Exercise 7

a.  

```{r}
# your code here - remove the hashtag! 
```

b.  YOUR ANSWER HERE

# Two categorical variables

## Exercise 8

a.  

```{r}
# your code here - remove the hashtag! 

# your code here - remove the hashtag! 
```

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

# One numeric & one categorical variable

## Exercise 9

a.  

```{r}
# your code here - remove the hashtag!

# your code here - remove the hashtag!

# your code here - remove the hashtag!
```

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

# More than two variables

## Exercise 10

a.  

```{r}
# your code here - remove the hashtag!
```

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

## Exercise 11

```{r}
# your code here - remove the hashtag!
```

YOUR ANSWER HERE

## Exercise 12

```{r}
# your code here - remove the hashtag!
```

YOUR ANSWER HERE

## Exercise 13

```{r}
# your code here - remove the hashtag!
```

YOUR ANSWER HERE

## Exercise 14

```{r}
# your code here - remove the hashtag!
```

YOUR ANSWER HERE

## Exercise 15

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

## BONUS

```{r}
# your code here - remove the hashtag!
```

YOUR ANSWER HERE
