---
title: "MLB Injuries- Introductory Time Series Analysis"
author: "YOUR NAME HERE"
format:
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: tango
editor_options:
  chunk_output_type: console
embed-resources: true
---

# Getting Started: MLB Injury Data

```{r, warning = FALSE, message = FALSE}
# Load the libraries
# Note you may need to install some of these packages
library(fpp3)
library(tidyverse)

# Load the data and convert to tsibble
injuries <- read_csv("mlb_injuries_monthly.csv") |> 
  mutate(Month = yearmonth(Month)) |> 
  as_tsibble(index = Month)
```

# Visualizing Time Series Data

## Exercise 1: Interpreting a Time Series Plot

```{r, echo = FALSE, warning = FALSE, message = FALSE}
autoplot(injuries, Count) +
  labs(title = "Monthly MLB Injury Counts",
       x = "Year",
       y = "Number of Injuries")
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

# Autocorrelation

## Exercise 2: Autocorrelation Plots

```{r, echo = FALSE, warning = FALSE, message = FALSE}
injuries |> 
  ACF(Count) |> 
  autoplot() +
  labs(title = "Autocorrelation of Monthly MLB Injury Counts",
       x = "Lag",
       y = "Autocorrelation")
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

# Decomposition

## Exercise 3: Changing Decomposition Types

a. 
```{r, warning=FALSE, message=FALSE, echo=FALSE}
injuries |> 
  model(classical_decomposition(Count)) |>
  components() |>
  autoplot() +
  labs(title = "Classical Additive Decomposition of Monthly MLB Injury Counts")
```

```{r}
#Your CODE HERE FOR CLASSICAL MULTIPLICATIVE DECOMPOSITION
```

YOUR ANSWER HERE

b. 
```{r, warning=FALSE, message=FALSE, echo=FALSE}
injuries |> 
  model(STL(Count ~ trend(window = 21)+ 
            season(window = "periodic"),
            robust = TRUE)) |>
  components() |>
  autoplot() +
  labs(title = "STL Additive Decomposition of Monthly MLB Injury Counts (Trend Window = 21)")
```

```{r}
#Your CODE HERE FOR STL Decomposition WITH SMALLER TREND WINDOW
```

YOUR ANSWER HERE

# Basic Forecasting Methods

## Exercise 4: Basic Forecasting Methods

a. YOUR ANSWER HERE

b. YOUR ANSWER HERE

c. YOUR ANSWER HERE

d. YOUR ANSWER HERE

e. YOUR ANSWER HERE

f. YOUR ANSWER HERE

# Evaluating Forecasting Models

## Exercise 5: Residual Analysis

```{r, message=FALSE, warning=FALSE, echo=FALSE}
snaive_mod <- injuries |>
  model(SNAIVE(Count ~ lag('year')))
  
snaive_mod |> 
  gg_tsresiduals()
```

a. YOUR ANSWER HERE

b. YOUR ANSWER HERE

c. YOUR ANSWER HERE

d. YOUR ANSWER HERE

```{r, echo=FALSE, eval=FALSE}
# This code is purely to help you later in the assignment
training <- injuries |> filter(year(Month) < 2018)
testing <- injuries |> filter(year(Month) >= 2018)

injury_fit <- training |>
  model(mean = MEAN(Count),
        naive = NAIVE(Count),
        snaive = SNAIVE(Count ~ lag('year')),
        drift = RW(Count ~ drift()))

injury_forecasts <- injury_fit |> 
  forecast(new_data = testing)

injury_forecasts |>
  autoplot(injuries, level = NULL) +
  labs(title = "Forecasting Methods for Monthly MLB Injury Counts")+
  guides(color = guide_legend(title = "Forecast"))

accuracy(injury_forecasts, testing)
```

# Forecasting

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# Fit the final model
final_fit <- injuries |> 
  model(SNAIVE(Count ~ lag('year')))

# Forecast the next 36 months
final_forecasts <- final_fit |> 
  forecast(h = 36)

# Plot the forecasts
final_forecasts |>
  autoplot(injuries) +
  labs(title = "Forecasting MLB Injuries")+
  guides(color = guide_legend(title = "Forecast"))
```

## Exercise 6: Forecasting MLB Injuries

a. YOUR ANSWER HERE

b. YOUR ANSWER HERE

c. YOUR ANSWER HERE

d. YOUR ANSWER HERE

# More Practice

## Challenge 1: Visualizing the Data

a. 
```{r, warning=FALSE, message=FALSE}
tj_surgeries <- read_csv("tj_surgeries_mlb_milb.csv") |> 
  as_tsibble(index = year)

# Your Code For Plotting Here
```

b. YOUR ANSWER HERE

c. YOUR ANSWER HERE

d. YOUR ANSWER HERE

e. 
```{r}
# Your Code For Plotting Acf Here
```

YOUR ANSWER HERE

f. YOUR ANSWER HERE

## Challenge 2: Simple Forecasting Methods

```{r}
training <- tj_surgeries |> filter(year <= 2011)
testing <- tj_surgeries |> filter(year > 2011)
```

a. 
```{r}
# Your Code For Fitting Models Here
```

b. 
```{r}
# Your Code For Plotting Forecasts Here
```

YOUR ANSWER HERE

c. 
```{r}
# Your Code For Calculating Accuracy Here
```

YOUR ANSWER HERE

d. 
```{r}
# Your Code for Fitting to the Full Data Here
# Your Code for Forecasting  and Plotting the Next 3 Years Here
```

YOUR ANSWER HERE

e. YOUR ANSWER HERE

f. YOUR ANSWER HERE


