---
title: "Masters - Single Mean Confidence Intervals"
author: "YOUR NAME HERE"
format:
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: tango
editor_options:
  chunk_output_type: console
---

# Getting Started: Masters 2023 Scores and Single Mean Confidence Intervals and Hypothesis Testing
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(BSDA)

masters_2023 <- read_csv("masters_2023.csv")
```

## Exercise 1: Data Structure

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

# Confidence Intervals for Single Means

## Exercise 2: T-Distribution Confidence Intervals

```{r}
amateurs_round1 <- masters_2023 |> 
  filter(round == 1, tour == "Amateur")
x_bar <- amateurs_round1 |> 
  summarise(mean(score)) |> 
  pull()
std_dev <- amateurs_round1 |> 
  summarize(sd(score)) |> 
  pull()
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

f.  YOUR ANSWER HERE

g.  YOUR ANSWER HERE

## Exercise 3: Z-Distribution Confidence Intervals

```{r}
pga_round1 <- masters_2023 |> 
  filter(round == 1, tour == "PGA")
x_bar <- pga_round1 |> 
  summarise(mean(score)) |> 
  pull()
sigma <- pga_round1 |> 
  summarise(sd(score)) |> 
  pull()
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

## Exercise 4: R for Confidence Intervals

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

# Hypothesis Testing with Single Mean Confidence Intervals

## Exercise 5: T-test for Single Mean Hypothesis Testing

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

## Exercise 6: Z-test for Single Mean Hypothesis Testing

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

# More Practice

## Challenge 1: Critical Values Knowledge Check

```{r}
liv_round3 <- masters_2023 |> 
  filter(round == 3, tour == "LIV")
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

## Challenge 2: P-Values Knowledge Check

```{r}
pro_round4 <- masters_2023 |> 
  filter(round == 4, tour == "PGA")
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

## Challenge 3: Confidence Interval Knowledge Check

```{r}
## Sample 1
amateur_round2 <- masters_2023 |> 
  filter(round == 2, tour == "Amateur")

## Sample 2
pga_rounds2_4 <- masters_2023 |> 
  filter(round %in% c(1, 2), tour == "PGA")
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE

f.  YOUR ANSWER HERE

g.  YOUR ANSWER HERE

## Challenge 4: Z vs T Large Sample Size

```{r}
pga_rounds1_2 <- masters_2023 |> 
  filter(round %in% c(1, 2), tour == "PGA")
```

a.  YOUR ANSWER HERE

b.  YOUR ANSWER HERE

c.  YOUR ANSWER HERE

d.  YOUR ANSWER HERE

e.  YOUR ANSWER HERE
